---
title: "CDA and EDA"
author: "Kevin Thompson and Stuart Miller"
date: "Last Updated June 27, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Setup

### Load Libraries

```{r libraries, echo=TRUE, message=FALSE}
library(tidyverse)
library(naniar)
library(reshape2)
library(pastecs)
library(maps)
library(mapproj)
```

### Load Data

Be sure to run `make` in `./analysis/data` first to build data file.

```{r load_data, echo=TRUE}
load('./data/.RData')
```

## Introduction

This is an analysis of a dataset containing beer and brewery data. The dataset contains 2410 beers from 558 breweries in 50 states of the United States and the District of Columbia. The goal was to explore the data set and look for interesting features of the data. Once the data was explored, the analysis focused on how international bitterness unit (IBU) and alcohol by volume (ABV) of beer are distributed over states and the relationship between IBU and ABV. The analysis also includes a discussion of the missing data and how the missing data impacts making insights from the dataset.

## Analysis Questions

The following questions were used to drive the data exploration.

**How many breweries are present in each state?**

```{r fig.width=15, fig.height=7, echo=TRUE}
# get counts of breweries per state
brewery_data %>% 
  group_by(State) %>% 
  summarise(breweries_per_state = length(State)) %>%
  ggplot(aes(x=reorder(State, -breweries_per_state), y=breweries_per_state)) +
  geom_col() +
  xlab('State') +
  ylab('Count') +
  ggtitle('Brewery Count per State')
```

Above is a plot of the number of breweries per state for each state and Washington DC.

About one-third of the states have more than 10 breweries, while the other two-thirds have less than 10. There are 47 breweries in Colorado and 39 in Califoria, which are the top two states by brewery count. The bottom three states by brewery count with just one brewery per state are North Dakota, South Dakota, and West Virginia. There is also only one brewery in Washington DC.

We next examine the data to ensure that data files were merged correctly.

```{r, echo=TRUE}
head(merged_data)
tail(merged_data)
```

Based on the first 6 and last 6 rows of the merged data, the merge appears to have been executed correctly. Note that this would only catch gross errors from the merge.


**Behavior of Missing Data**

```{r}
merged_data %>%
  miss_var_summary()
```

The table below, shows the number of missing values (`NA`) for `IBU`, `ABV`, `Style`, and `ShortStyle`. A large portion of the `IBU` values are missing, which would affect insightes related to `IBU`. The number of missing values for `ABV` is less significant, but caution should be taken when working with this column. The other two columns are missing less than 1% of the rows.

| Column Name   | Number of Missing Values| Percent of Values Missing|
|:-------------:|------------------------:|-------------------------:|
| IBU           | 1005                    |41.70%                    |
| ABV           | 62                      |2.57%                     |
| Style         | 5                       |0.21%                     |
| ShortStyle    | 5                       |0.21%                     |

Because only 2.57% of ABV is missing and there is no indication of a pattern for the missingness, we are comfortable assuming we can recover a consistent estimate of the mean for Alcohol-By-Volume after list-wise deletion of missing data. The missingness is quite low between states and there doesn't appear to be any distinct geographical pattern. 


```{r maps, echo=TRUE}
missing.by.state <- merged_data %>% 
  group_by(region) %>% 
  summarise(ABV_missing=prop_miss(ABV) * 100,
            IBU_missing=prop_miss(IBU) * 100)

states <- map_data("state")
map.df <- merge(states, missing.by.state, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]

ggplot(map.df, aes(x=long,y=lat,group=group, fill=ABV_missing))+
  geom_polygon()+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+
  coord_map()+
  labs(title = 'Percent Missing ABV by State')
```

We turn to the question of missingness in IBU, which is a more complicated topic. 

```{r, warnings=FALSE, messages=FALSE}
ggplot(map.df, aes(x=long,y=lat,group=group, fill=IBU_missing))+
  geom_polygon()+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+
  coord_map() +
  labs(title = 'Percent Missing IBU by State')



#missingness by style
nabular_data %>%
  group_by(ShortStyle) %>%
  summarize(perc_miss_IBU = prop_miss(IBU) * 100,
            num_miss_IBU = n_miss(IBU)) %>%
  filter(num_miss_IBU > 15) %>%
  ggplot(aes(x=reorder(ShortStyle, desc(perc_miss_IBU)), y=perc_miss_IBU)) +
  geom_col() +
  labs(title="Missingness of IBU by Beer Style", x = "Beer Style", y = "% Missing IBU         Values")
```



There is clearly a relationship between the beer style, location, and missingness of IBU, so IBU cannot be missing completely at random; therefore, listwise deletion is not acceptable. We will have to look into the question of whether we can build a decent imputation model for IBU at some point in the future. Depending on the nature of the missingness, most parameters for the distribution of IBU may not be recoverable. 

**Comparison of Alcohol Content and Bitterness between States**

```{r, echo=TRUE}
# calc summary stats for ABV and IBU
summary_df <- merged_data %>% 
  remove_missing(na.rm=TRUE) %>%
  group_by(region) %>% 
  summarise(MeanABV=mean(ABV), MaxABV=max(ABV), MinABV=min(ABV), MedianABV=median(ABV), StdABV=sd(ABV),
            MeanIBU=mean(IBU), MaxIBU=max(IBU), MinABV=min(IBU), MedianIBU=median(IBU), StdIBU=sd(IBU))

# build data for mapping
states <- map_data("state")
map.df <- merge(states, summary_df, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
```


```{r fig.width=15, fig.height=7, echo=TRUE}
merged_data %>%
  remove_missing(na.rm=TRUE) %>%
  group_by(State) %>%
  summarise(MedianIBU = median(IBU)) %>%
  ggplot(aes(x=reorder(State, -MedianIBU), y=MedianIBU)) +
  geom_col() +
  labs(title="Median IBU by State", x="State", y="Median IBU")
```

The plot above shows the median IBU values per state. The medians range from 61 to 19 with the highest median from Maine and the lowest median from Wisconsin.

```{r, echo=TRUE}
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=MedianIBU))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+
  coord_map()+
  labs(title = 'Median IBU by State')
```

The heat map of median IBU by state above shows some regional clustering of median IBU. The median of the upper west states is on the higher side of IBU, the median of the midwest is near the lower end of IBU medians, and the deep south is on the top end of IBU medians. 

```{r fig.width=15, fig.height=7, echo=TRUE}
merged_data %>%
  remove_missing(na.rm=TRUE) %>%
  group_by(State) %>%
  summarize(MedianABV = median(ABV)) %>%
  ggplot(aes(x=reorder(State, -MedianABV), y=MedianABV)) +
  geom_col() +
  labs(title="Median ABV by State", x="State", y="Median ABV")

```

The plot above shows the median ABV values per state. The medians range from 6.25% to 4% with the highest median from Maine and the lowest median from Utah.

```{r, echo=TRUE}
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=MedianABV))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+
  coord_map()+
  labs(title = 'Median ABV by State')
```

The heat map of median ABV by state above shows some regional clustering of median ABV. The deep south, some central south states, and some states on the east coast are near the top end of the ABV median scale. The medians of the midwest though the west coast are more moderate on the ABV median scale.


**Which states have the maximum alcoholic (`ABV`) and bitter (`IBU`) beer?**

```{r, fig.width=15, fig.height=7, echo=TRUE}
#Max IBUs
merged_data %>%
  remove_missing(na.rm=TRUE) %>%
  group_by(State) %>%
  summarize(MaxIBU = max(IBU)) %>%
  ggplot(aes(x=reorder(State, -MaxIBU), y=MaxIBU)) +
  geom_col() +
  labs(title="Maximum IBU by State", x="State", y="Maximum IBU")

#Max ABVs 
merged_data %>%
  remove_missing(na.rm=TRUE) %>%
  group_by(State) %>%
  summarize(MaxABV = max(ABV)) %>%
  ggplot(aes(x=reorder(State, -MaxABV), y=MaxABV)) +
  geom_col() +
  labs(title="Maximum ABV by State", x="State", y="Maximum ABV")
```

The state producing the most bitter beer is Oregon with an IBU of 138. The state producing the most alcoholic beer is Colorado with an ABV of 12.8%.


**Summary Statistics for the ABV variable.**

```{r, echo=TRUE}
merged_data %>%
  summary()
```

The following table summerizes the summary statistics for ABV.

| Statistic           | Value |
|--------------------:|------------------------:|
| Min                 | 0.1%                    |
| 1st Quartile        | 5%                      |
| Mean                | 5.98%                   |
| Median              | 5.60%                   |
| 3rd Quartile        | 6.7%                    |
| Max                 | 12.8%                   | 
| standard deviation  | 1.35 percentage points  |
| range               | 12.7 percentage points  |

The bulk of the data has an ABV between 5% and 6.7%. The mean and median are fairly different on this scale, likely indicating skew or influence of outliers in the distribution of this variable.

**Is there an apparent relationship between the bitterness of the beer and its alcoholic content?**

```{r scatterplot/regression, echo=TRUE}

# fits best-fit line to scatter plot
merged_data %>%
  remove_missing(na.rm=TRUE) %>%
  ggplot(aes(x=IBU, y=ABV)) +
  geom_point(alpha=0.1) +
  geom_smooth(method="lm") + 
  labs(title="ABU vs IBU")

# model output
lm(ABV ~ IBU, data=merged_data) %>%
  summary()

# effect size
source("background_code/effect_size.R")
full_model <- lm(ABV ~ IBU, data=merged_data)
reduced_model <- lm(ABV ~ 1, data=merged_data)
get_effect_size(full_model, reduced_model)
```

The plot above is a scatter plot of ABV vs IBU. A simple linear regression yields strong evidence of a linear relationship (p-value < 0.0001). We find that a marginal increase in IBU increases ABV by 0.03%. We compute an effect size of 0.817 using Cohen's F-squared, which means that the proportion of the variance accounted for above and beyond the sample mean of ABV is very significant. 


## Conclusion

### Results of the Exploration and Analysis

* There appears to be geographoc clustering of the median IBU and median ABV of states.
* Most beers have a moderate ABV value (5% - 6.7%).
* There is evidence of a linear relationship between ABV and IBU.

### inferential Scope of Analysis**

The details behind the sampling process are unkwown, so it is unclear whether we can generalize our results to a wider population. We also cannot infer causation because there is insufficient theoretical knowledge to make good causal assumptions, especially wit hthe absence of randomization. The missingness of the data poses a problem as well. While the missingness of ABV lends itself to listwise deletion, the missingness of IBU is clearly not completely at random. At best, the missingness of IBU is caused primarily by its location and beer style, but further examination is needed to determine whether an imputation model can recover a consistent estimate of key parameters, like the population mean. If a consistent estimate cannot be recovered, then sampling processes and data governance must be re-examined to obtain better results. 


