---
title: "CDA and EDA"
author: "Kevin Thompson and Stuart Miller"
date: "June 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Setup

### Load Libraries

```{r libraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(pastecs)
```

### Load Data

Be sure to run `make` in `./analysis/data` first.

```{r}
load('./data/.RData')
```
## Introduction

TODO: Add an introduciton to the project

## Analysis Questions

TODO: Add short header here

### Question 1

```{r fig.width=15, fig.height=7}
# get counts of breweries per state
brewerysPerState <- brewery_data %>% 
  group_by(State) %>% 
  summarise(Brewery_name = length(State))
# order the result
brewerysPerState <- brewerysPerState[order(-brewerysPerState$Brewery_name), ]
# better name for column
names(brewerysPerState)[2] <- 'BreweryCount'
# plot count vs state (bar)
ggplot(data=brewerysPerState, aes(x = reorder(State, -BreweryCount), y = BreweryCount)) +
  geom_bar(stat = 'identity') +
  xlab('State') +
  ylab('Count') +
  ggtitle('Brewery Count per State')
```

TODO: Add description

### Question 2

TODO: (Kevin) Add description

Print 'head' and 'tail' of merged data to verify merge data looks correct.

```{r}
head(merged_data)
tail(merged_data)
```

TODO: Add description

### Question 3

```{r}
nas.in.merged.data <- colSums(is.na(merged_data))
nas.in.merged.data
```

The table below, shows the number of missing values (`NA`) for `IBU` and `ABV`. No other columns are missing values. A large portion of the `IBU` values are missing, which would affect insightes related to `IBU`. The number of missing values for `ABV` is less significant, but caution should be taken when working with this column.

| Column Name   | Number of Missing Values| Percent of Values Missing|
|:-------------:|------------------------:|-------------------------:|
| IBU           | 1005                    |41.70%                    |
| ABV           | 62                      |2.57%                     |


### Question 4

Calculate median alcohol content and internation bitterness unit for each state.

```{r}
# calc summary stats for IBU
IBUByState <- merged_data %>% 
  filter(!is.na(IBU)) %>%
  group_by(State) %>% 
  summarise(Mean=mean(IBU), Max=max(IBU), Min=min(IBU), Median=median(IBU), Std=sd(IBU))

# calc summary stats for ABV
ABVByState <- merged_data %>% 
  filter(!is.na(ABV)) %>%
  group_by(State) %>% 
  summarise(Mean=mean(ABV), Max=max(ABV), Min=min(ABV), Median=median(ABV), Std=sd(ABV))

# melt ByState data.frames to view stats side-by-side
IBUByState.long<-melt(IBUByState, id.vars = 'State')
ABVByState.long<-melt(ABVByState, id.vars = 'State')
```

```{r fig.width=15, fig.height=7}
ggplot(IBUByState.long[IBUByState.long$variable %in% c('Median'), ], 
       aes(x=reorder(State, -value), y=value, fill = variable)) +
    geom_bar(stat='identity', position='dodge') +
  ggtitle('Median of IBU per State') +
  xlab('State') + ylab('Median Value of ABV')
```

TODO: Add description of this graph

```{r fig.width=15, fig.height=7}
ggplot(ABVByState.long[ABVByState.long$variable %in% c('Median'), ], 
       aes(x=reorder(State, -value), y=value, fill = variable)) +
    geom_bar(stat='identity', position='dodge') +
  ggtitle('Median of ABV per State') +
  xlab('State') + ylab('Median Value of ABV')
```

TODO: Add description of this graph

### Question 5

```{r}
#IBUByState.long[IBUByState.long$variable %in% c('Max','Min'), ]
maxIBU <- IBUByState.long %>% filter(variable == 'Max') %>% arrange(-value)
paste('State with maximum IBU beer:', as.character(maxIBU[1,'State']), sep = '')
maxABV <- ABVByState.long %>% filter(variable == 'Max') %>% arrange(-value)
paste('State with maximum ABV beer:', as.character(maxABV[1,'State']), sep = '')
```

TODO: Add description

### Question 6

TODO: clean up this table

```{r}
res <- stat.desc(data.frame(merged_data$ABV))
round(res, 4)
```

TODO: Add description

### Question 7

TODO: Add code for this; scatter plot with lm? Check for statistical significance of model?

TODO: Add description

## Conclusion

TODO: Add conclusion to project; 
